---
title: "github_poster_figures"
author: "C T"
date: "2025-1-24"
output: html_document
---


# PACKAGES ----------------------------------------------------------------

```{r}
library(lubridate)
library(dplyr)
library(purrr)
library(tidyverse)
library(knitr)
library(sf)
library(mapview)
library(stringr)
library(tidytext)
```

Set filepath to data folder
```{r}
# celee's
# data_path <- "FRESES_DATA"

# maggie's 
data_path <- "data"
```

# DATA ---------------------------------------------------------------------

```{r}
# read in analysis-ready monarch dataset
df_final <- read.csv(paste0(data_path, "/IMMP_MLMP_final.csv"))

# read in region shapefile
cons_units <- st_read(paste0(data_path, "/Monarch_Butterfly_US_Conservation_Units/Monarch_Butterfly_US_Conservation_Units.shp"))
```


# POSTER PLOTS/GRAPHS

## Figure 2
Super hacky shapefile of FL
Note: I got this county shp from the Census TIGER site (it may be down now...) https://www.census.gov/geographies/mapping-files/time-series/geo/tiger-line-file.html 
```{r}
# read in shapefile of counties -> limit to FL
fl <- read_sf(paste0(data_path, "/cb_2018_us_county_5m/cb_2018_us_county_5m.shp")) %>% 
  filter(STATEFP == 12) %>%          # keep Florida
  st_union() %>%                     # combine all the counties into 1 polygon
  st_as_sf() %>%                     
  st_transform(4326)                 # convert to WGS84

# ok - that didn't quite get all of FL though. Manually modify the limits
bbox["ymax"] <- 30.35  
bbox["xmin"] <- -87.36
bbox["ymin"] <- 24.5 
bbox["xmax"] <- -80

# Convert modified bbox to sf object
bbox_sf <- st_as_sf(st_as_sfc(bbox), crs = st_crs(cons_units))

# Add bbox to original FL outline
fl_full <- st_union(fl, bbox_sf)
```


Map of site ID locations over our 3 regions
```{r}
#Format map data into our 3 desired regions (W,N,S)
cons_units <- cons_units %>%
  mutate(region = case_when(
    str_starts(NAME, "North") ~ "north",     
    str_starts(NAME, "South") ~ "south",   
    str_starts(NAME, "West") ~ "west" )) %>%
  st_transform(4326) 

# consolidate sub-region polygons into regions
cons_units <- cons_units %>%
  group_by(region) %>% 
  summarize(geometry = st_as_sf(st_union(geometry)))

# drop FL from the cons_units map
cons_units <- st_difference(cons_units, fl_full) 

# Convert dataframe to shapefile and add region - NOTE: the 250 or so sites without coordinates were lost
sf_all_region = df_final %>%
  filter(!is.na(longitude) & !is.na(latitude)) %>%    
  st_as_sf(coords=c("longitude", "latitude"), crs=4326, remove = FALSE) %>% 
  st_join(cons_units_no_fl, left = TRUE) 

# Create Map
mapview(cons_units_no_fl, zcol = "region", legend = TRUE, layer.name = "Region") +  
  mapview(sf_all_region, color = "#04033F", alpha = 1, cex = 3, legend = FALSE)
```

## Figure 3
Avg. Monarch Presence by Milkweed Spp and Region
```{r}
# Calculate the average monarch presence per species per region
monarch_presence_by_region <- df_final %>%
  group_by(region, mwspecies, nativity) %>%  # Group by region, species, and nativity
  summarise(avg_presence = mean(monpresence, na.rm = TRUE), .groups = "drop")  # Compute average presence

# Drop A. Sullivantii from south bc it's always absence=0: 
monarch_presence_by_region <- monarch_presence_by_region %>% 
  filter(!(region=="south" & mwspecies == "Asclepias sullivantii"))

# Abbreviate species names using scientific name format (Genus initial + species)
monarch_presence_by_region <- monarch_presence_by_region %>%
  mutate(abbreviated_species = str_replace_all(mwspecies, 
                                               "^(\\w)\\w+ (\\w+)",  # Regex to match genus and species
                                               "\\1. \\2"))  # Abbreviate to Genus initial. Species

# Convert nativity to a factor with meaningful labels
monarch_presence_by_region$species_type <- factor(monarch_presence_by_region$nativity,
                                                  levels = c(0, 1),
                                                  labels = c("Non-native", "Native"))

# Plot the results
p2 <- ggplot(monarch_presence_by_region, 
        aes(x = reorder_within(abbreviated_species, avg_presence, region), 
           y = avg_presence, 
           fill = species_type))  +
  geom_bar(stat = "identity", width = 0.7) +  # Use fill based on species type
  coord_flip() +
  #facet_wrap(~ region, scales = "free_y", ncol = 1) +  # Stack graphs vertically by setting ncol=1
  facet_grid(region ~ ., scales = "free_y", space = "free") + # makes the bar widths uniform, but titles are stuck to the side now...
  labs(
    title = "Average Monarch Presence by Milkweed Species and Region",
    x = "Milkweed Species",
    y = "Avg. Monarch Presence"
  ) +
  scale_fill_manual(values = c("#E69F00", "#56B4E9"),  # Set colors for Non-native and Native
                    labels = c("Non-native", "Native")) +  # Labels for the legend
  guides(fill = guide_legend(title = NULL)) +  
  theme_minimal() +
  theme(
    plot.title = element_text(size = 25, face = "bold", hjust = 0.5),  # Title size and centered (hjust = 0.5)
    axis.title = element_text(size = 25),       # Axis titles size
    axis.text = element_text(size = 15),        # Axis labels size
    axis.text.y = element_text(size = 15),      # Increase size of species names
    strip.text = element_text(size = 25),       # Size for the region labels (facets)
    legend.text = element_text(size = 20) ,     # Legend text size
    legend.position = "bottom",                 # Moves the legend to the bottom
    legend.direction = "horizontal",            # Arranges the legend items horizontally
    panel.spacing = unit(1, "lines")            # Increase space between facets
  ) + 
  scale_x_reordered() # remove suffixes

# print p2
p2 
```

## Figure 3b 
TOTAL Monarchs by Milkweed Spp and Region
```{r}
# Calculate the average monarch presence per species per region
monarch_presence_by_region <- df_final %>%
  group_by(region, mwspecies, nativity) %>%  # Group by region, species, and nativity
  summarise(tot_monarchs = sum(monarchcount, na.rm = TRUE), .groups = "drop")  # Compute average presence

# Drop A. Sullivantii from south bc it's always absence=0: 
monarch_presence_by_region <- monarch_presence_by_region %>% 
  filter(!(region=="south" & mwspecies == "Asclepias sullivantii"))

# Abbreviate species names using scientific name format (Genus initial + species)
monarch_presence_by_region <- monarch_presence_by_region %>%
  mutate(abbreviated_species = str_replace_all(mwspecies, 
                                               "^(\\w)\\w+ (\\w+)",  # Regex to match genus and species
                                               "\\1. \\2"))  # Abbreviate to Genus initial. Species

# Convert nativity to a factor with meaningful labels
monarch_presence_by_region$species_type <- factor(monarch_presence_by_region$nativity,
                                                  levels = c(0, 1),
                                                  labels = c("Non-native", "Native"))

# Plot the results
p2 <- ggplot(monarch_presence_by_region, 
        aes(x = reorder_within(abbreviated_species, tot_monarchs, region), 
           y = tot_monarchs, 
           fill = species_type))  +
  geom_bar(stat = "identity", width = 0.7) +  # Use fill based on species type
  coord_flip() +
  #facet_wrap(~ region, scales = "free_y", ncol = 1) +  # Stack graphs vertically by setting ncol=1
  facet_grid(region ~ ., scales = "free_y", space = "free") + # makes the bar widths uniform, but titles are stuck to the side now...
  labs(
    title = "Total Monarchs by Milkweed \n Species and Region",
    x = "Milkweed Species",
    y = "Total Monarch Count"
  ) +
  scale_fill_manual(values = c("#E69F00", "#56B4E9"),  # Set colors for Non-native and Native
                    labels = c("Non-native", "Native")) +  # Labels for the legend
  guides(fill = guide_legend(title = NULL)) +  
  theme_minimal() +
  theme(
    plot.title = element_text(size = 25, face = "bold", hjust = 0.5),  # Title size and centered (hjust = 0.5)
    axis.title = element_text(size = 25),       # Axis titles size
    axis.text = element_text(size = 15),        # Axis labels size
    axis.text.y = element_text(size = 15),      # Increase size of species names
    strip.text = element_text(size = 25),       # Size for the region labels (facets)
    legend.text = element_text(size = 20) ,     # Legend text size
    legend.position = "bottom",                 # Moves the legend to the bottom
    legend.direction = "horizontal",            # Arranges the legend items horizontally
    panel.spacing = unit(1, "lines")            # Increase space between facets
  ) + 
  scale_x_reordered() # remove suffixes

# print p2
p2 
```

## Figure 3c
TOTAL Monarchs / TOTAL Milkweed Plants -  by Milkweed Spp and Region
```{r}
# Calculate the average monarch presence per species per region
monarch_presence_by_region <- df_final %>%
  group_by(region, mwspecies, nativity) %>%  # Group by region, species, and nativity
  summarise(tot_monarchs = sum(monarchcount)/sum(mwcount), na.rm = TRUE, .groups = "drop")  # Compute average presence

# Drop A. Sullivantii from south bc it's always absence=0: 
monarch_presence_by_region <- monarch_presence_by_region %>% 
  filter(!(region=="south" & mwspecies == "Asclepias sullivantii"))

# Abbreviate species names using scientific name format (Genus initial + species)
monarch_presence_by_region <- monarch_presence_by_region %>%
  mutate(abbreviated_species = str_replace_all(mwspecies, 
                                               "^(\\w)\\w+ (\\w+)",  # Regex to match genus and species
                                               "\\1. \\2"))  # Abbreviate to Genus initial. Species

# Convert nativity to a factor with meaningful labels
monarch_presence_by_region$species_type <- factor(monarch_presence_by_region$nativity,
                                                  levels = c(0, 1),
                                                  labels = c("Non-native", "Native"))

# Plot the results
p2 <- ggplot(monarch_presence_by_region, 
        aes(x = reorder_within(abbreviated_species, tot_monarchs, region), 
           y = tot_monarchs, 
           fill = species_type))  +
  geom_bar(stat = "identity", width = 0.7) +  # Use fill based on species type
  coord_flip() +
  #facet_wrap(~ region, scales = "free_y", ncol = 1) +  # Stack graphs vertically by setting ncol=1
  facet_grid(region ~ ., scales = "free_y", space = "free") + # makes the bar widths uniform, but titles are stuck to the side now...
  labs(
    title = "Total Monarchs per Plant, \n by Milkweed Species and Region",
    x = "Milkweed Species",
    y = "Monarch Density"
  ) +
  scale_fill_manual(values = c("#E69F00", "#56B4E9"),  # Set colors for Non-native and Native
                    labels = c("Non-native", "Native")) +  # Labels for the legend
  guides(fill = guide_legend(title = NULL)) +  
  theme_minimal() +
  theme(
    plot.title = element_text(size = 25, face = "bold", hjust = 0.5),  # Title size and centered (hjust = 0.5)
    axis.title = element_text(size = 25),       # Axis titles size
    axis.text = element_text(size = 15),        # Axis labels size
    axis.text.y = element_text(size = 15),      # Increase size of species names
    strip.text = element_text(size = 25),       # Size for the region labels (facets)
    legend.text = element_text(size = 20) ,     # Legend text size
    legend.position = "bottom",                 # Moves the legend to the bottom
    legend.direction = "horizontal",            # Arranges the legend items horizontally
    panel.spacing = unit(1, "lines")            # Increase space between facets
  ) + 
  scale_x_reordered() # remove suffixes

# print p2
p2 
```

## Figure 4a
Seasonal Dynamics of Monarch Presence on the Top 5 Milkweed Species within each Region
```{r}
# Calculate `monthly_avg` by averaging monarch presence per species per month in each region
monthly_avg <- df_final %>%
  group_by(region, mwspecies, month) %>%
  summarise(avg_presence = mean(monpresence, na.rm = TRUE), .groups = "drop")

# List of top 5 species for each region
n_species <- c("Asclepias syriaca", "Asclepias sullivantii", "Asclepias incarnata", "Asclepias exaltata", "Asclepias curassavica")
s_species <- c("Asclepias tuberosa", "Asclepias syriaca", "Asclepias incarnata", "Asclepias curassavica", "Asclepias asperula")
w_species <- c("Asclepias speciosa", "Asclepias fascicularis", "Asclepias eriocarpa", "Asclepias curassavica", "Asclepias cordifolia")

# Combine all species
all_species <- unique(c(n_species, s_species, w_species))

# Filter the data for only the species you're interested in
monthly_avg_filtered <- monthly_avg %>%
  filter(mwspecies %in% all_species)

# Add a new column for abbreviated species names
monthly_avg_filtered <- monthly_avg_filtered %>%
  mutate(
    abbreviated_species = str_replace_all(mwspecies, 
                                         "^(\\w)\\w+ (\\w+)",  # Regex to match genus and species
                                         "\\1. \\2")  # Abbreviate to Genus initial. Species
  )

# Ensure Month is numeric (if it's not already)
monthly_avg_filtered <- monthly_avg_filtered %>%
  mutate(month = as.numeric(month))

# Plot the average monarch presence by month
ggplot(monthly_avg_filtered, aes(x = month, y = avg_presence, color = abbreviated_species)) +
  geom_smooth(
    method = "loess", 
    se = TRUE, 
    linewidth = 1,      # Adjust the line width of the smooth line
    size = 0.5,         # Adjust the boundary size of the CI (makes it thinner)
    fill = "lightgrey", # Change CI color
    alpha = 0.3         # Adjust transparency of the CI (0 is fully transparent, 1 is fully opaque)
  ) + 
  scale_x_continuous(
    breaks = 1:12, 
    labels = substr(month.name, 1, 1)  # Use first letter of each month
  ) +  # Set x-axis labels to the first letter of the month
  scale_color_manual(
    values = c(
      # South (Vibrant Green)
      "A. syriaca" = "#4D9B3B",  # vibrant green 
      "A. sullivantii" = "#6BBF39", # bright green 
      "A. incarnata" = "#A0D94A",  # yellowish green 
      "A. curassavica" = "red",  # red for curassavica 
      "A. exaltata" = "#99B85F",  # olive green
     
      # North (Purple)
      "A. syriaca" = "#6A3D9C",  # deep purple
      "A. incarnata" = "#9F61D3",  # lavender purple
      "A. tuberosa" = "#9B6FCC",  # soft lavender
      "A. asperula" = "#8B58B3",  # muted purple
      
      # West (Blue)
      "A. fascicularis" = "#006DCC", # deep vibrant blue
      "A. speciosa" = "#1E90FF",  # dodger blue
      "A. eriocarpa" = "#4682B4",  # steel blue
      "A. cordifolia" = "#5B8D9D"  # medium blue-gray
    ),
    name = "Milkweed Species"  # Legend title
  ) +
  labs(
    title = "Seasonal Dynamics of Larvae Populations on Popular Milkweed Species",
    x = "Month",
    y = "Average Monarch Presence",
    color = "Milkweed Species"
  ) +
  facet_wrap(~ region) +  # Add region as a facet to separate by region
  theme_classic() +
  theme(
    plot.title = element_text(size = 22, face = "bold", hjust = 1),       # Title size
    axis.title = element_text(size = 22),                      # Axis titles size
    axis.text = element_text(size = 18),                       # Axis labels size
    axis.text.x = element_text(size = 18),                     # Axis numbers size (increase text size for x-axis)
    axis.text.y = element_text(size = 18),                     # Axis numbers size
    legend.title = element_text(size = 20, face = "bold"),     # Legend title size
    legend.text = element_text(size = 20),                     # Legend labels size
    strip.text = element_text(size = 15, face = "bold")        # Region labels (facet labels) size and style
  )

```

## Figure 4b
Seasonal Dynamics of Monarch Presence on the Top 5 Milkweed Species within each Region - BAR PLOT
```{r}
# Calculate `monthlyseasonal_avg_avg` by averaging monarch presence per species per SEASON in each region
seasonal_avg <- df_final %>%
  mutate(season = case_when(
    julday >= 355 | julday < 80 ~ "winter",
    julday >= 80 & julday < 172 ~ "spring",
    julday >= 172 & julday < 265 ~ "summer",
    julday >= 265 & julday < 355 ~ "fall"
  )) %>% 
  group_by(region, nativity, season) %>%
  # use sum instead of mean if you wanna look at total counts
  summarise(avg_presence = mean(monpresence, na.rm = TRUE), .groups = "drop") %>%  
  mutate(nativity = factor(nativity,
                           levels = c(0, 1),
                           labels = c("Non-native", "Native")))

# Plot the average monarch presence by month
ggplot(seasonal_avg, aes(x = season, y = avg_presence, fill = nativity)) +
  geom_bar(stat="identity", position="dodge")  +
  facet_wrap(~ region) + 
  scale_fill_manual(values = c("#E69F00", "#56B4E9"),  # Set colors for Non-native and Native
                    labels = c("Non-native", "Native")) +  # Labels for the legend
  guides(fill = guide_legend(title = NULL)) +  
  theme_minimal() +
  labs(
    title = "Seasonal Dynamics of Larvae Populations on Popular Milkweed Species",
    x = "Season",
    y = "Avg. Monarch Presence"  
    ) +
  theme(
    plot.title = element_text(size = 22, face = "bold", hjust = 1),       # Title size
    axis.title = element_text(size = 22),                      # Axis titles size
    axis.text = element_text(size = 18),                       # Axis labels size
    axis.text.x = element_text(size = 18),                     # Axis numbers size (increase text size for x-axis)
    axis.text.y = element_text(size = 18),                     # Axis numbers size
    legend.title = element_text(size = 20, face = "bold"),     # Legend title size
    legend.text = element_text(size = 20),                     # Legend labels size
    strip.text = element_text(size = 15, face = "bold")        # Region labels (facet labels) size and style
  )

```


