---
title: "modeling"
author: "C T"
date: "2025-01-17"
output: html_document
---

Packages

```{r}
#load relevant libraries
x <- c("Matrix", "MASS", "plotrix", "lme4",  "lmerTest", "lmtest", "gridExtra",
       "DHARMa", "sjPlot", "ggeffects", "performance", "MuMIn", "emmeans", "tidyverse")
lapply(x, library, character.only = TRUE)
rm(x)
```

Data

```{r}
# Read in CSV
df_final <- read_csv("data/IMMP_MLMP_final.csv")

# Reset column types (must do this *EVERY TIME* you read in the file)
df_final <- df_final %>%
  mutate(monpresence = as.factor(monpresence),
         mwspecies = as.factor(mwspecies),
         siteid = as.factor(siteid),
         year = as.numeric(year),
         region = as.factor(region),
         nativity = as.factor(nativity),
         month = as.factor(month)
         )
```

Break into regional datasets

```{r}
df_south <- subset(df_final, region == "south")
df_north <- subset(df_final, region == "north")
df_west <- subset(df_final, region == "west")
```

## Nativity model

In the South, the nativity model does not converge when year is a fixed effect

```{r}
# SOUTH 1: nativity + year + month + (1 | siteid) 
# DOES NOT CONVERGE
model_nativity_s1 <- glmer(
  monpresence ~ nativity + year + month + (1 | siteid), 
  family = binomial(link = "logit"),
  data = df_south, control = glmerControl(optimizer="bobyqa"))

summary(model_nativity_s1) # AIC = 5711.8


# SOUTH 2: try with year as random effect now 
# CONVERGES
model_nativity_s2 <- glmer(
  monpresence ~ nativity + month + (1 | year) + (1 | siteid), 
  family = binomial(link = "logit"),
  data = df_south, control = glmerControl(optimizer="bobyqa",
                                          optCtrl=list(maxfun=2500))) 

summary(model_nativity_s2) # AIC = 5674.1


# SOUTH 3: try with month as a random effect 
# DOES NOT CONVERGE
model_nativity_s3 <- glmer(
  monpresence ~ nativity +  year + (1 | month) +  + (1 | siteid), 
  family = binomial(link = "logit"),
  data = df_south, control = glmerControl(optimizer="bobyqa"))

summary(model_nativity_s3) # # AIC = 5730.5


# *SOUTH 4: try with month AND year as a random effects
# CONVERGES
model_nativity_s4 <- glmer(
  monpresence ~ nativity + (1 | month) + (1 | year) + (1 | siteid), 
  family = binomial(link = "logit"),
  data = df_south, control = glmerControl(optimizer="bobyqa"))

summary(model_nativity_s4) # AIC = 5692.8
```

Model 4 was best, for the south. Let's run it for north and west too.

```{r}
# north
model_nativity_n4 <- glmer(
  monpresence ~ nativity + (1 | month) + (1 | year) + (1 | siteid), 
  family = binomial(link = "logit"),
  data = df_north, control = glmerControl(optimizer="bobyqa"))

# west
model_nativity_w4 <- glmer(
  monpresence ~ nativity + (1 | month) + (1 | year) + (1 | siteid), 
  family = binomial(link = "logit"),
  data = df_west, control = glmerControl(optimizer="bobyqa"))
```

Use emmeans to interpret results:

```{r}
# USE EMMEANS TO INTERPRET RESULTS
emmeans(model_nativity_s4, pairwise ~ nativity, type = "response")
emmeans(model_nativity_n4, pairwise ~ nativity, type = "response")
emmeans(model_nativity_w4, pairwise ~ nativity, type = "response")
```

-   In the South, the probability for monarch presence is (0.18) on native and (0.65) on nonnative spp. Odds for non-natives having monarchs are 8.4 times higher than natives

-   In the North, the odds radio is 1.6 and barely significant at 0.01

-   In the West, the odds ratio is 4

## Milkweed species model

Again, the mwspecies model does not converge when year is a fixed effect (in the south). Month is insignificant and actually lowers AIC, whether used as a random or fixed effect.

```{r}
# SOUTH 1: species + year + (1 | siteid) 
# DOES NOT CONVERGE
model_mwspecies_s1 <- glmer(
  monpresence ~ mwspecies + year + (1 | siteid), 
  family = binomial(link = "logit"),
  data = df_south, control = glmerControl(optimizer="bobyqa"))

summary(model_mwspecies_s1) # AIC = 5702.1

# SOUTH 2: now try year as random effect
# CONVERGES
model_mwspecies_s2 <- glmer(
  monpresence ~ mwspecies + (1 | year) + (1 | siteid), 
  family = binomial(link = "logit"),
  data = df_south, control = glmerControl(optimizer="bobyqa"))

summary(model_mwspecies_s2) # AIC = 5674.2

# SOUTH 3: add month
# CONVERGES
model_mwspecies_s3 <- glmer(
  monpresence ~ mwspecies + month + (1 | year) + (1 | siteid), 
  family = binomial(link = "logit"),
  data = df_south, control = glmerControl(optimizer="bobyqa"))

summary(model_mwspecies_s3) # AIC = 5402.3

# SOUTH 4: try month as a random effect
# CONVERGES
model_mwspecies_s4 <- glmer(
  monpresence ~ mwspecies + (1 | month) + (1 | year) + (1 | siteid), 
  family = binomial(link = "logit"),
  data = df_south, control = glmerControl(optimizer="bobyqa"))

summary(model_mwspecies_s4) # AIC = 5421.6
```

Model 2 was best, for the south. Let's run it for north and west too.

```{r}
# north
model_mwspecies_n2 <- glmer(
  monpresence ~ mwspecies + (1 | year) + (1 | siteid), 
  family = binomial(link = "logit"),
  data = df_north, control = glmerControl(optimizer="bobyqa"))

# west
model_mwspecies_w2 <- glmer(
  monpresence ~ mwspecies + (1 | year) + (1 | siteid), 
  family = binomial(link = "logit"),
  data = df_west, control = glmerControl(optimizer="bobyqa"))
```

```{r}
# USE EMMEANS TO INTERPRET RESULTS
emmeans_mwsp_s <- emmeans(model_mwspecies_s2, pairwise ~ mwspecies, type = "response")
emmeans_mwsp_n <- emmeans(model_mwspecies_n2, pairwise ~ mwspecies, type = "response")
emmeans_mwsp_w <- emmeans(model_mwspecies_w2, pairwise ~ mwspecies, type = "response")
```

-   In the north, species like Asclepias curassavica and Asclepias syriaca have relatively high probabilities of monarch presence

-   In the west, curassavica = high probability of monarch presence, followed by fascicularis

Plot emmeans alltogether:

```{r}
emmeans_mwsp_s_df <- as.data.frame(emmeans_mwsp_s$emmeans) %>% mutate(region="south")
emmeans_mwsp_w_df <- as.data.frame(emmeans_mwsp_w$emmeans) %>% mutate(region="west")
emmeans_mwsp_n_df <- as.data.frame(emmeans_mwsp_n$emmeans) %>% mutate(region="north")
emmmeans_mwsp_all <- bind_rows(emmeans_mwsp_s_df, emmeans_mwsp_w_df, emmeans_mwsp_n_df)

# Create a unique list of x-axis labels
x_labels <- levels(factor(emmmeans_mwsp_all$mwspecies))

# Define color vector based on x-axis labels
label_colors <- ifelse(x_labels == "Asclepias curassavica", "red", "black")

# Set order of x-axis labels
order <- emmmeans_mwsp_all %>%
  group_by(mwspecies) %>%
  summarize(mean_prob = mean(prob, na.rm = TRUE)) %>%
  arrange(desc(mean_prob)) %>%
  pull(mwspecies)

# Plot with proper alignment of label colors
emmmeans_mwsp_all %>%
  mutate(mwspecies = factor(mwspecies, levels = order)) %>%
  ggplot(aes(x = mwspecies, y = prob, color = region, ymin = asymp.LCL, ymax = asymp.UCL)) +
  geom_point(size = 3) +
  geom_errorbar(width = 0.2) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, color = label_colors)
  )
```

Or plot emmeans separately...

```{r}

# Initialize an empty list to store the plots
plot_list <- list()

# List of your data frames
emmeans_list <- list(emmeans_mwsp_s_df, emmeans_mwsp_w_df, emmeans_mwsp_n_df)

# Loop through each data frame
for(i in emmeans_list) {
  
  # Set order of x-axis labels
  order <- i %>%
    arrange(desc(prob)) %>%
    pull(mwspecies)
  
  # Create a unique list of x-axis labels
  x_labels <- levels(factor(i$mwspecies))
  
  # Create plot and add to the list
  plot_list[[length(plot_list) + 1]] <- ggplot(i, aes(x = mwspecies,
                                                      y = prob, 
                                                      color = region, 
                                                      ymin = asymp.LCL,
                                                      ymax = asymp.UCL)) +
    geom_point(size = 3) + 
    geom_errorbar(width = 0.2) + 
    theme_minimal(base_size = 14) + 
    scale_color_manual(values = c("north" = "red", "south" = "green", "west" = "blue")) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "none"
    ) +
    scale_x_discrete(limits = order) +  
    labs(title = paste("Region:", unique(i$region)))  
}

# Display the plots in a grid
grid.arrange(grobs = plot_list, ncol = 3)
```
