---
title: "modeling"
author: "C T"
date: "2025-01-17"
output: html_document
---

Packages

```{r}
#load relevant libraries
x <- c("Matrix", "MASS", "plotrix", "lme4",  "lmerTest", "lmtest", "gridExtra",
       "DHARMa", "sjPlot", "ggeffects", "performance", "MuMIn", "emmeans", "tidyverse", "grid")
lapply(x, library, character.only = TRUE)
rm(x)
```

Data

```{r}
# Read in CSV
df_final <- read_csv("../data/IMMP_MLMP_final.csv")

# Reset column types (must do this *EVERY TIME* you read in the file)
df_final <- df_final %>%
  mutate(monpresence = as.factor(monpresence),
         mwspecies = as.factor(mwspecies),
         siteid = as.factor(siteid),
         year = as.numeric(year),
         region = as.factor(region),
         nativity = as.factor(nativity),
         month = as.factor(month)
         ) 

# (for figures later) add a new column for abbreviated species names
df_final <- df_final %>%
  mutate(
    abb_spp = str_replace_all(mwspecies, 
                                         "^(\\w)\\w+ (\\w+)",  # Regex to match genus and species
                                         "\\1. \\2")  # Abbreviate to Genus initial. Species
  )

```



Break into regional datasets

```{r}
df_south <- subset(df_final, region == "south")
df_north <- subset(df_final, region == "north")
df_west <- subset(df_final, region == "west")
```

## General Summary stats

count the number of species and surveys we examined in our filtered dataset
```{r}
# number of milkweed species
num_species <- n_distinct(df_final$abb_spp)
# number of surveys
num_visits <- n_distinct(df_final$visitid)

# print results
cat("Number of species:", num_species, "\n")
cat("Number of surveys:", num_visits, "\n")
```





## Nativity models

In the South, the nativity model does not converge when year is a fixed effect

```{r}
# SOUTH 1: nativity + year + month + (1 | siteid) 
# DOES NOT CONVERGE
model_nativity_s1 <- glmer(
  monpresence ~ nativity + year + month + (1 | siteid), 
  family = binomial(link = "logit"),
  data = df_south, control = glmerControl(optimizer="bobyqa"))

summary(model_nativity_s1) # AIC = 5711.8


# SOUTH 2: try with year as random effect now 
# CONVERGES
model_nativity_s2 <- glmer(
  monpresence ~ nativity + month + (1 | year) + (1 | siteid), 
  family = binomial(link = "logit"),
  data = df_south, control = glmerControl(optimizer="bobyqa",
                                          optCtrl=list(maxfun=2500))) 

summary(model_nativity_s2) # AIC = 5674.1


# SOUTH 3: try with month as a random effect 
# DOES NOT CONVERGE
model_nativity_s3 <- glmer(
  monpresence ~ nativity +  year + (1 | month) +  + (1 | siteid), 
  family = binomial(link = "logit"),
  data = df_south, control = glmerControl(optimizer="bobyqa"))

summary(model_nativity_s3) # # AIC = 5730.5


# *SOUTH 4: try with month AND year as a random effects
# CONVERGES
model_nativity_s4 <- glmer(
  monpresence ~ nativity + (1 | month) + (1 | year) + (1 | siteid), 
  family = binomial(link = "logit"),
  data = df_south, control = glmerControl(optimizer="bobyqa"))

summary(model_nativity_s4) # AIC = 5692.8
```

Model 4 was best, for the south. Let's run it for north and west too.

```{r}
# north
model_nativity_n4 <- glmer(
  monpresence ~ nativity + (1 | month) + (1 | year) + (1 | siteid), 
  family = binomial(link = "logit"),
  data = df_north, control = glmerControl(optimizer="bobyqa"))

# west
model_nativity_w4 <- glmer(
  monpresence ~ nativity + (1 | month) + (1 | year) + (1 | siteid), 
  family = binomial(link = "logit"),
  data = df_west, control = glmerControl(optimizer="bobyqa"))
```

Use emmeans to interpret results:

```{r}
# USE EMMEANS TO INTERPRET RESULTS
emmeans(model_nativity_s4, pairwise ~ nativity, type = "response")
emmeans(model_nativity_n4, pairwise ~ nativity, type = "response")
emmeans(model_nativity_w4, pairwise ~ nativity, type = "response")
```

-   In the South, the probability for monarch presence is (0.18) on native and (0.64) on nonnative spp. Odds for non-natives having monarchs are 8.3 times higher than natives; p-value = <0.0001

-   In the North, the odds radio is 1.6; p value = 0.0188

-   In the West, the odds ratio is 3.9; p value =  <0.0001


(poster figure 6) Graph ALL nativity emmeans results (best N, S, W models) into a nice figure
```{r}
#  create objects for eemeans from each model
   # west:
  emmeans_natvity_w4 <- emmeans(model_nativity_w4, pairwise ~ nativity, type = "response")
    # south
  emmeans_nativity_s4 <- emmeans(model_nativity_s4, pairwise ~ nativity, type = "response")
    # north
  emmeans_nativity_n4 <- emmeans(model_nativity_n4, pairwise ~ nativity, type = "response")

# Convert to dataframes
    #South
s_nativity_emmeans <- as.data.frame(emmeans_nativity_s4)
s_nativity_emmeans$region <- "south"
    # west
w_nativity_emmeans <- as.data.frame(emmeans_natvity_w4)
w_nativity_emmeans$region <- "west"
    # north
n_nativity_emmeans <- as.data.frame(emmeans_nativity_n4)
n_nativity_emmeans$region <- "north"
 

# make x-axis display "Native" and "Non-native" instead of "0" and "1" for each region
n_nativity_emmeans$nativity <- factor(n_nativity_emmeans$nativity,
                                        levels = c(0, 1), 
                                        labels = c("Non-native", "Native"))
s_nativity_emmeans$nativity <- factor(s_nativity_emmeans$nativity,
                                        levels = c(0, 1), 
                                        labels = c("Non-native", "Native"))

w_nativity_emmeans$nativity <- factor(w_nativity_emmeans$nativity,
                                        levels = c(0, 1), 
                                        labels = c("Non-native", "Native"))
# Remove rows where nativity is NA for each region
n_nativity_emmeans <- n_nativity_emmeans %>% filter(!is.na(nativity))
s_nativity_emmeans <- s_nativity_emmeans %>% filter(!is.na(nativity))
w_nativity_emmeans <- w_nativity_emmeans %>% filter(!is.na(nativity))


# Initialize an empty list to store the plots
plot_list <- list()

# List of your data frames
emmeans_list <- list(n_nativity_emmeans, s_nativity_emmeans, w_nativity_emmeans)

# Loop through each data frame
for(i in emmeans_list) {
  
  # Set order of x-axis labels
  order <- i %>%
    arrange(desc(prob)) %>%
    pull(nativity)
  
  # Create a unique list of x-axis labels
  x_labels <- levels(factor(i$nativity))
  
  # Create plot and add to the list
  plot_list[[length(plot_list) + 1]] <- ggplot(i, aes(x = nativity,
                                                      y = prob, 
                                                      color = region, 
                                                      ymin = asymp.LCL,
                                                      ymax = asymp.UCL)) +
    geom_point(size = 7) + 
    geom_errorbar(width = 0.4, size = 2) + 
    theme_minimal(base_size = 30) + 
    scale_color_manual(values = c("north" = "#0072B2", "south" = "#E69F00", "west" = "#009E73")) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      axis.title = element_blank(),     ## MAGGIE DID THIs
      legend.position = "none",
      plot.margin = margin(20,10,10,60),
      plot.background = element_rect(fill = "#FFF5E1"),  # background for the whole plot
      axis.line = element_line(color = "black", size = 1)
    ) +
    scale_x_discrete(limits = order) +  
    scale_y_continuous(limits = c(0,1), breaks=c(0,1)) + ## MAGGIE DID THIs
    labs(subtitle = paste("Region:", unique(i$region)))  
}

# Display the plots in a grid
grid.arrange(
  grobs = plot_list,
  ncol = 3,
  top = textGrob("Modeled Probability of Monarch Presense by Natvity and Region", gp = gpar(fontsize = 28, fontface = "bold"))
  )
```




## Milkweed species model

Again, the mwspecies model does not converge when year is a fixed effect (in the south). Month is insignificant and actually lowers AIC, whether used as a random or fixed effect.

```{r}
# SOUTH 1: species + year + (1 | siteid) 
# DOES NOT CONVERGE
model_mwspecies_s1 <- glmer(
  monpresence ~ mwspecies + year + (1 | siteid), 
  family = binomial(link = "logit"),
  data = df_south, control = glmerControl(optimizer="bobyqa"))

summary(model_mwspecies_s1) # AIC = 5702.1

# SOUTH 2: now try year as random effect
# CONVERGES
model_mwspecies_s2 <- glmer(
  monpresence ~ abb_spp + (1 | year) + (1 | siteid), 
  family = binomial(link = "logit"),
  data = df_south, control = glmerControl(optimizer="bobyqa"))

summary(model_mwspecies_s2) # AIC = 5674.2

# SOUTH 3: add month
# CONVERGES
model_mwspecies_s3 <- glmer(
  monpresence ~ mwspecies + month + (1 | year) + (1 | siteid), 
  family = binomial(link = "logit"),
  data = df_south, control = glmerControl(optimizer="bobyqa"))

summary(model_mwspecies_s3) # AIC = 5402.3

# SOUTH 4: try month as a random effect
# CONVERGES
model_mwspecies_s4 <- glmer(
  monpresence ~ mwspecies + (1 | month) + (1 | year) + (1 | siteid), 
  family = binomial(link = "logit"),
  data = df_south, control = glmerControl(optimizer="bobyqa"))

summary(model_mwspecies_s4) # AIC = 5421.6
```


Model 2 was best, for the south. Let's run it for north and west too.

```{r}
# north
model_mwspecies_n2 <- glmer(
  monpresence ~ abb_spp + (1 | year) + (1 | siteid), 
  family = binomial(link = "logit"),
  data = df_north, control = glmerControl(optimizer="bobyqa"))

# west
model_mwspecies_w2 <- glmer(
  monpresence ~ abb_spp + (1 | year) + (1 | siteid), 
  family = binomial(link = "logit"),
  data = df_west, control = glmerControl(optimizer="bobyqa"))
```

```{r}
# USE EMMEANS TO INTERPRET RESULTS
emmeans(model_mwspecies_s2, pairwise ~ abb_spp, type = "response")
emmeans(model_mwspecies_n2, pairwise ~ abb_spp, type = "response")
emmeans(model_mwspecies_w2, pairwise ~ abb_spp, type = "response")
```
Results: 

-   In the north, species like Asclepias curassavica and Asclepias syriaca have relatively high probabilities of monarch presence

-   In the west, curassavica = high probability of monarch presence, followed by fascicularis

- p values:
 ---> north:
        A currasavica = 0.0187
        A. incarnata = 0.6867 
  ---> south =
        A. curaasavica = 0.0002
        A. incarnata = 0.0452
  ---> west =
      A. currasavica =  <.0001
      A. eriocarpa = 0.0157


```{r}
# Make species emmeans dataframes
emmeans_mwsp_s <- emmeans(model_mwspecies_s2, pairwise ~ abb_spp, type = "response")
emmeans_mwsp_n <- emmeans(model_mwspecies_n2, pairwise ~ abb_spp, type = "response")
emmeans_mwsp_w <- emmeans(model_mwspecies_w2, pairwise ~ abb_spp, type = "response")
```



(Poster figure 5) plot all milkweed species emmeans (N,S,W) into a nice figure ...
```{r}

# Initialize an empty list to store the plots
plot_list <- list()

# List of your data frames
emmeans_list <- list(emmeans_mwsp_n_df, emmeans_mwsp_s_df, emmeans_mwsp_w_df)

# Loop through each data frame
for(i in emmeans_list) {
  
  # Set order of x-axis labels
  order <- i %>%
    arrange(desc(prob)) %>%
    pull(abb_spp)
  
  # Create a unique list of x-axis labels
  x_labels <- levels(factor(i$abb_spp))
  
  # Create plot and add to the list
  plot_list[[length(plot_list) + 1]] <- ggplot(i, aes(x = abb_spp,
                                                      y = prob, 
                                                      color = region, 
                                                      ymin = asymp.LCL,
                                                      ymax = asymp.UCL)) +
    geom_point(size = 7) + 
    geom_errorbar(width = 0.4, size = 2) + 
    theme_minimal(base_size = 30) + 
    scale_color_manual(values = c("north" = "#0072B2", "south" = "#E69F00", "west" = "#009E73")) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      axis.title = element_blank(),     ## MAGGIE DID THIs
      legend.position = "none",
      plot.margin = margin(20,10,10,70),
      plot.background = element_rect(fill = "#FFF5E1"),  # background for the whole plot
      panel.border = element_blank(),   # Remove panel borders
      axis.line = element_line(color = "black", size = 1)
    ) +
    scale_x_discrete(limits = order) +  
    scale_y_continuous(limits = c(0,1), breaks=c(0,1)) + ## MAGGIE DID THIs
    labs(subtitle = paste("Region:", unique(i$region)))  
}

# Display the plots in a grid
grid.arrange(
  grobs = plot_list,
  ncol = 3,
  top = textGrob("Modeled Probability of Monarch Presense by Species and Region", gp = gpar(fontsize = 28, fontface = "bold"))
  )
```

Plot milkweed species emmeans alltogether (NOT by region):
```{r}
emmeans_mwsp_s_df <- as.data.frame(emmeans_mwsp_s$emmeans) %>% mutate(region="south")
emmeans_mwsp_w_df <- as.data.frame(emmeans_mwsp_w$emmeans) %>% mutate(region="west")
emmeans_mwsp_n_df <- as.data.frame(emmeans_mwsp_n$emmeans) %>% mutate(region="north")
emmmeans_mwsp_all <- bind_rows(emmeans_mwsp_s_df, emmeans_mwsp_w_df, emmeans_mwsp_n_df)

# Create a unique list of x-axis labels
x_labels <- levels(factor(emmmeans_mwsp_all$abb_spp))

# Define color vector based on x-axis labels
label_colors <- ifelse(x_labels == "Asclepias curassavica", "darkred", "darkgreen")

# Set order of x-axis labels
order <- emmmeans_mwsp_all %>%
  group_by(abb_spp) %>%
  summarize(mean_prob = mean(prob, na.rm = TRUE)) %>%
  arrange(desc(mean_prob)) %>%
  pull(abb_spp)

# Plot with proper alignment of label colors
emmmeans_mwsp_all %>%
  mutate(abb_spp = factor(abb_spp, levels = order)) %>%
  ggplot(aes(x = abb_spp, y = prob, color = region, ymin = asymp.LCL, ymax = asymp.UCL)) +
  geom_point(size = 3) +
  geom_errorbar(width = 0.2) +
  theme_minimal(base_size = 30) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, color = label_colors)
  ) +
  labs(title = "Modelled Probability of Monarch Presence")
```





#----------------------------------------------------------------------------------------
# UNIQUE SITEIDS  & VISIT IDS PER REGION -----------------------------------------------------------
#----------------------------------------------------------------------------------------

# **SOUTH**
```{r}
sites_sp_s <- df_south %>%
  group_by(mwspecies) %>%
  summarise(
    unique_siteids = n_distinct(siteid),
    unique_visitids = n_distinct(visitid),
    .groups = "drop"
  ) %>%
  arrange(desc(unique_siteids)) 

print(sites_sp_s)
```

#sites
```{r}
ggplot(sites_sp_s, aes(x = reorder(mwspecies, unique_siteids), y = unique_siteids)) +
  geom_bar(stat = "identity", fill = "green") +
  coord_flip() +  # Flip coordinates for better readability
  labs(
    title = "SOUTH - SITES",
    x = "Milkweed Species",
    y = "# Unique Sites"
  ) +
  theme_minimal()
```


# visits
```{r}
ggplot(sites_sp_s, aes(x = reorder(mwspecies, unique_visitids), y = unique_visitids)) +
  geom_bar(stat = "identity", fill = "green") +
  coord_flip() +  # Flip coordinates for better readability
  labs(
    title = "SOUTH - SURVEYS",
    x = "Milkweed Species",
    y = "Number of Visits/Surveys"
  ) +
  theme_minimal()
```


# _____
# **NORTH** 
```{r}
sites_sp_n <- df_north %>%
  group_by(mwspecies) %>%
  summarise(
    unique_siteids = n_distinct(siteid),
    unique_visitids = n_distinct(visitid),
    .groups = "drop"
  ) %>%
  arrange(desc(unique_siteids)) 

print(sites_sp_n)
```

#sites
```{r}
ggplot(sites_sp_n, aes(x = reorder(mwspecies, unique_siteids), y = unique_siteids)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  coord_flip() +  # Flip coordinates for better readability
  labs(
    title = "NORTH - SITES",
    x = "Milkweed Species",
    y = "# Unique Sites"
  ) +
  theme_minimal()
```

# visits
```{r}
ggplot(visitid_counts_n, aes(x = reorder(mwspecies, total_visitids), y = total_visitids)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  coord_flip() +  # Flip coordinates for better readability
  labs(
    title = "NORTH - SURVEYS",
    x = "Milkweed Species",
    y = "Number of Visits/Surveys"
  ) +
  theme_minimal()
```


#________
# **WEST**
```{r}
sites_sp_w <- df_west %>%
  group_by(mwspecies) %>%
  summarise(
    unique_siteids = n_distinct(siteid),
    unique_visitids = n_distinct(visitid),
    .groups = "drop"
  ) %>%
  arrange(desc(unique_siteids)) 

print(sites_sp_w)
```

#sites
```{r}
ggplot(sites_sp_w, aes(x = reorder(mwspecies, unique_siteids), y = unique_siteids)) +
  geom_bar(stat = "identity", fill = "orange") +
  coord_flip() +  # Flip coordinates for better readability
  labs(
    title = "WEST - SITES",
    x = "Milkweed Species",
    y = "# Unique Sites"
  ) +
  theme_minimal()
```

#visits/surveys
```{r}
ggplot(sites_sp_w, aes(x = reorder(mwspecies, unique_visitids), y = unique_visitids)) +
  geom_bar(stat = "identity", fill = "orange") +
  coord_flip() +  # Flip coordinates for better readability
  labs(
    title = "WEST - SURVEYS",
    x = "Milkweed Species",
    y = "Number of Visits/Surveys"
  ) +
  theme_minimal()
```


